KSPHPDDM and PCHPDDM: extending PETSc with advanced Krylov methods and robust multilevel overlapping Schwarz preconditioners

### 2020-07-24 article.tex
Fix labelling in figure caption
Style fixes

\documentclass[3p,11pt]{elsarticle}

\usepackage{lineno,hyperref}
% \modulolinenumbers[5]

\journal{Computers \& Mathematics with Applications}
\newcommand\col{:}
\newcommand{\pk}[1]{\texttt{#1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\usepackage{colortbl,booktabs}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox,amssymb}
\usepackage{cleveref}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

\begin{frontmatter}

\title{KSPHPDDM and PCHPDDM: extending PETS\MakeLowercase{c} with advanced Krylov methods and robust multilevel Overlapping Schwarz preconditioners}

\author[firstaddress]{Pierre Jolivet\corref{correspondingauthor}}
\cortext[correspondingauthor]{Corresponding author}
\ead{pierre.jolivet@enseeiht.fr}

\author[secondaryaddress]{Jose E. Roman}
\ead{jroman@dsic.upv.es}

\author[thirdaddress]{Stefano Zampini}
\ead{stefano.zampini@kaust.edu.sa}

\address[firstaddress]{CNRS, IRIT--ENSEEIHT, Toulouse, France}
\address[secondaryaddress]{Universitat Polit\`ecnica de Val\`encia, Val\`encia, Spain}
\address[thirdaddress]{King Abdullah University of Science and Technology, Thuwal, Saudi Arabia}

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced in order to offer
    end-users robust overlapping Schwarz preconditioners and advanced Krylov
    methods featuring recycling and the ability to deal with multiple right-hand sides.
    The flexibility of the implementation %, which for example may deal with various matrix formats,
    is showcased and explained with minimalist, easy-to-run, and reproducible examples,
    to ease the integration of these algorithms into more advanced frameworks.
    The examples provided cover applications from
    %simulate the linear or nonlinear solves of typical
    %applications in fracture mechanics, linear stability analysis or more
    %broadly-speaking
    eigenanalysis, elasticity, combustion, and
    electromagnetism.
\end{abstract}

@@

\end{frontmatter}

\linenumbers

@@

Fruits of the proposed enhancement are discussed for applications from
SLEPc~\cite{hernandez2005ssf}, the Scalable Library for Eigenvalue Problem
computations, an add-on library that extends PETSc with classes for
linear and nonlinear eigenvalue problems, as well as with tools to facilitate
the use of shift-and-invert spectral transformations.

@@

Because Krylov methods may need long-term recurrences to mitigate round-off
errors during the generation of subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the volume of
global communications needed when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. On the one hand, PETSc implements the loose GMRES~\cite{baker2005technique}~(LGMRES) and the
deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES) as
\pk{KSPLGMRES} and \pk{KSPDGMRES}. However, neither handle variable
preconditioning, unlike \pk{KSPFGMRES} or \pk{KSPGCR} which respectively
implement the flexible GMRES~\cite{saad1993flexible}~(FGMRES) and the
generalized conjugate residual method~\cite{eisenstat1983variational}~(GCR).
Furtheremore, \pk{KSPDGMRES} does not support complex arithmetic.  Moreover,
extending the recycling capabilities of such methods to sequence of linear
systems with smoothly varying coefficient matrices and right-hand sides is not
trivial.  On the other hand, HPDDM offers support for the generalized conjugate
residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR), which does not suffer from the
aforementioned limitations.

@@

For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}s:
\begin{itemize}
    \item \pk{MATAIJ}: standard sequential or parallel sparse matrix, based on compressed sparse row format;
    \item \pk{MATSEQBAIJ}: block sparse matrix, based on block compressed sparse row format;
    \item \pk{MATSEQSBAIJ}: symmetric block sparse matrix stored in upper triangular form;
    \item \pk{MATSHELL}: user-defined matrix;
    \item \pk{MATNEST}: block-defined matrix with nested submatrices;
    \item \pk{MATAIJCUSPARSE}: sequential or parallel sparse matrix, offloaded to a NVIDIA GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}
The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    %\item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: embedded Krylov method;
    \item \pk{PCMAT}: matrix multiplication;
    %\item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    %\item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}: hierarchical matrices~\cite{10.1145/3232850,ambartsumyan2020hierarchical}; 
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like
preconditioners~(\cref{it:hpddm,it:asm,it:bjacobi}) is that they most often
rely on exact or inexact factorizations~(\cref{it:lu,it:ilu}) as subdomain
solvers.  In these cases, it is possible to access the so-called factored
matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$,
%$\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to
take advantage of blocked forward eliminations and backward substitutions from
the various factorization packages interfaced with PETSc. In the case of a
sparse matrix, this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL PARDISO or
CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense matrix:
ScaLAPACK and Elemental~\cite{poulson2013elemental} are supported. Future work
will consider extending the multigrid framework \pk{PCMG}, as well as other
preconditioning classes to increase arithmetic intensity of the preconditioner
application phase for blocks of right-hand sides.

@@

    \item in PETSc with no preconditioning and no restart GMRES($\infty$) \pk{KSPGMRES};

@@

In all of the cases, convergence is declared when the initial unpreconditioned
residual is reduced by 10 orders of magnitude.

@@

processes and can be launched using the following command lines. \\[-4pt]

@@

        \caption{Number of iterations needed to converge for various

@@

        the number of iterations needed by \pk{KSPHPDDM} match with the respective

@@

        xlabel={Number of right-hand sides ($N$)},

@@

The mini-app is then used to benchmark intranode performance of matrix--matrix
multiplications, either with multiple OpenMP threads, or by offloading the
operation to a NVIDIA GPU and using cuSPARSE~\cite{cusparse-web-page} as
interfaced in PETSc.

@@

    \subsubsection{Linear stability analysis}
In this section, we apply recycling Krylov methods in the context of linear
stability analysis and consider the solution of the following generalized eigenvalue problem:
\begin{equation}\label{eq:lsa}
    J(q_b)x = \lambda \begin{bmatrix}M & 0 \\ 0 & 0 \end{bmatrix} x,
\end{equation}
where $J$ is the Jacobian of the incompressible steady-state Navier--Stokes equation
evaluated using a given base flow $q_b = \begin{bmatrix}u_b \\ p_b\end{bmatrix}$ and
$M$ is the discretization of the mass matrix on the space of velocities.
The mini-app employed for the numerical results is built on top of
FreeFEM~\cite{hecht2012new} and it is available at
\url{https://github.com/prj-/moulin2019al}.
Interested
readers are referred to~\cite{moulin2018al} for more details and for larger
runs.

@@

The Krylov--Schur algorithm converges in six outer iterations (restarts), with a total of 46 inner solves~\cref{eq:ks}.
Recycling of Krylov subspaces with \pk{KSPHPDDM} provides algorithmic speedup by lowering the number of iterations per inner solve,
as reported in~\cref{fig:comparison-slepc}, from 80 with \pk{KSPFGMRES} to 45.
A similar speedup can be observed in terms of runtime, with the time spent for
the solution of the eigenvalue problem being reduced from \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{25.1}{\minute}
to \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{17.5}{\minute}.
These timings, and all that follow, have been obtained on
Irène, a system composed of \pgfmathprintnumber[assume math mode=true]{1656}
nodes with two 24-core Intel Xeon Platinum 8168 clocked at
\SI{2.7}{\giga\hertz}.
\pgfplotstableread{
sys hpddm fgmres
1   77    77
2   54    83
3   53    83
4   52    81
5   52    81
6   51    78
7   51    79
8   51    80
9   51    80
10  51    77
11  51    79
12  51    79
13  51    79
14  53    77
15  52    79
16  49    78
17  51    77
18  51    79
19  52    79
20  52    79
21  51    77
22  51    78
23  51    79
24  53    80
25  54    79
26  52    79
27  49    75
28  51    78
29  51    78
30  51    77
31  50    78
32  55    81
33  56    79
34  54    80
35  53    80
36  53    78
37  50    78
38  52    79
39  53    77
40  52    77
41  54    80
42  53    80
43  55    80
44  53    79
45  52    77
46  53    80
}\loadedtableSLEPc
    \begin{figure}[!t]
\centering \begin{tikzpicture}
    \begin{axis}[xlabel={Krylov--Schur system index},ylabel={\# of
        iterations},xmin=1,xmax=46,ymin=0,ymajorgrids,
        legend columns=1,enlarge x limits=0.05, height=5.5cm,
        legend style={font=\small,at={(0.98,0.02)}, legend cell align=left,
        anchor=south east}]
\addplot[very thick,last]     table[x expr=\thisrowno{0},y expr=\thisrowno{2}, col sep=space] {\loadedtableSLEPc};\label{pgfplots:m0}
\addplot[very thick,myred,densely dashed]       table[x expr=\thisrowno{0},y expr=\thisrowno{1}, col sep=space] {\loadedtableSLEPc};\label{pgfplots:m1}
\legend{{\pk{KSPFGMRES}},{\pk{KSPHPDDM}}}
\end{axis}
\end{tikzpicture}
        \caption{Number of inner iterations for systems~\cref{eq:ks} with~\ref{pgfplots:m1} or without~\ref{pgfplots:m0} recycling.\label{fig:comparison-slepc}}
    \end{figure}

\subsubsection{Blocking inside SLEPc\label{sec:lobpcg}}
The implementations of the locally optimal block preconditioned conjugate
gradient~\cite{knyazev2001toward}~(LOBPCG) and the contour integral spectrum
slicing method~\cite{sakurai2003projection}~(CISS) in SLEPc were previously not
using blocking when applying the preconditioner or solving linear systems. In
version 3.14.0, these solvers have been adapted to employ \pk{KSPMatSolve} and
\pk{PCMatApply}.  In the case of LOBPCG, which is a purely blocked method, all
steps were already implemented as block operations, except the applications of
the preconditioner, which undermined the benefits of blocking.
%With the new
%implementation, the gains are striking, as reported in~\cref{sec:lobpcg}.
The performance of LOBPCG implemented in SLEPc as \pk{EPSLOBPCG} is studied using three different preconditioners:
\begin{itemize}
    \item \pk{PCASM}: one-level overlapping Schwarz method with one level of overlap and exact Cholesky factorizations in each subdomain;
    \item \pk{PCHPDDM}: multilevel overlapping Schwarz method, see~\cref{sec:PCHPDDM};
    \item \pk{PCGAMG}: algebraic multigrid method~\cite{adams2004ultra}.
\end{itemize}
More details on these solvers are given in the next section where focus is put
on preconditioning rather than Krylov methods.  We note here that only
\pk{PCASM} and \pk{PCHPDDM} currently handle blocking, see the
list~\cref{sec:interface-ksp}, and that \pk{PCHPDDM} and \pk{PCGAMG} have lower
contraction factors than the cheaper alternative \pk{PCASM}.

@@

The script available at \url{https://github.com/prj-/jolivet2020petsc} may be
used to solve the same eigenproblem using \pk{EPSCISS} by providing the
additional command line argument \pk{-eps\_type ciss}.
However, this eigensolver is mostly suited for finding interior eigenpairs.
For fairness, its performance is not reported, but the options from the script
can be readily used for problems where \pk{EPSCISS} is an appropriate solver.
We finally note that blocking support is also available and tested in SLEPc for \pk{EPSSUBSPACE}
since version 3.14.0.

@@

HPDDM implements the same one-level overlapping Schwarz methods as PETSc as
well as optimized Schwarz methods~\cite{gander2006optimized}, which may be used
in PETSc with \pk{PCSetModifySubMatrices}, see~\cite{gander2019new}.  It also provides support for
defining robust two-level methods using the generalized eigenvalue problem on
the overlap (GenEO) framework, for finite
elements~\cite{spillane2011robust,spillane2013abstract} and boundary
elements~\cite{marchand2019two}. Results in this work present the first
high-level access to GenEO from PETSc and its composable solver
infrastructure~\cite{brown2012composable}. We note that GenEO has been
recently implemented in other libraries as well~\cite{butler2020high}.

@@

conditions. In the case of the one-dimensional Poisson

@@

\caption{Notation and basics of overlapping Schwarz methods.\label{fig:poisson}}

@@

For performance, it is better to compute the same number of
eigenvectors on each process, since in this case, the coarse operator will be
assembled using symmetric or general blocked matrix formats. With
usually large block sizes, this has consequent impacts on memory, floating-point, and message passing performances.
% When applying the preconditioner,
%\texttt{MPI\_Gather} and \texttt{MPI\_Scatter} operations are used to move
%between levels, instead of the variable-size alternatives \texttt{MPI\_Gatherv}
%and \texttt{MPI\_Scatterv}, that could exhibit worse performance.
In addition,
standard eigensolvers such as \pk{EPSKRYLOVSCHUR} from SLEPc, cannot inexpensively
evaluate the exact number of eigenpairs in a given interval of $\mathbb{R}^+$.
Instead, it is often more efficient to provide an upper bound on the number of
eigenpairs.

@@

decomposition methods, \pk{PCNN} and \pk{PCBDDC} must be supplied with a \pk{MatIS}

@@

it is possible to obtain the Dirichlet operators

@@

overlapping Schwarz methods are known for converging slowly when there is
no overlap. The proposed methodology is not strictly equivalent to the

@@

    \varepsilon_{ij}(u) = \frac{1}{2} \left(\frac{\partial u_i}{\partial x_i} + \frac{\partial u_j}{\partial x_j}\right)\!, \text{ } \mu = \frac{E}{2(1+\nu)}, \text{ and }\lambda = \frac{E\nu}{1-2\nu}.

@@

    \subfloat[\parbox{5cm}{Number of inner coarse iterations for various values of $M$, the number of second-level subdomains.}\label{fig:inexact}]{

@@

configuration with 256 second-level subdomains~\ref{pgfplots:p-256}, requires
64 seconds to setup and 36 seconds for the solution of the linear system.

>>>>

% SIAM Article Template
\documentclass[review,onefignum,onetabnum]{siamart190516}

% Information that is shared between the article and the supplement
% (title and author information, macros, packages, etc.) goes into
% ex_shared.tex. If there is no supplement, this file can be included
% directly.

\input{shared}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={KSPHPDDM and PCHPDDM: Extending PETSc with Robust Overlapping Schwarz Preconditioners and Advanced Krylov Methods},
  pdfauthor={P. Jolivet, J. E. Roman, and S. Zampini}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

% \externaldocument{supplement}

% FundRef data to be entered by SIAM
%<funding-group specific-use="FundRef">
%<award-group>
%<funding-source>
%<named-content content-type="funder-name">
%</named-content>
%<named-content content-type="funder-identifier">
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>
\newcommand\col{:}
\newcommand{\pk}[1]{\texttt{#1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\usepackage{colortbl,booktabs}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

\maketitle

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    end-users robust overlapping Schwarz preconditioners and advanced Krylov
    methods featuring recycling and the ability to deal with multiple right-hand sides.
    The flexibility of the implementation %, which for example may deal with various matrix formats,
    is showcased and explained with minimalist, easy-to-run, and reproducible examples,
    to ease the integration of these algorithms into more advanced frameworks.
    The examples provided cover applications from
    %simulate the linear or nonlinear solves of typical
    %applications in fracture mechanics, linear stability analysis or more
    %broadly-speaking
    eigenanalysis, elasticity, combustion, and
    electromagnetism.
\end{abstract}

@@

% REQUIRED
\begin{AMS}
  35Q68, 65Y05
\end{AMS}

@@

Fruits of the proposed enhancement are discussed for applications from
SLEPc~\cite{hernandez2005ssf}, the Scalable Library for Eigenvalue Problem
computations, an add-on library that extends PETSc with several classes for
linear and nonlinear eigenvalue problems, as well as with tools to facilitate
the use of shift-and-invert spectral transformations.

@@

Because Krylov methods may need long-term recurrences to mitigate round-off
errors during the generation of subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the volume of
global communications needed when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. On the one hand, PETSc implements the Loose GMRES~\cite{baker2005technique}~(LGMRES) and the
Deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES) as
\pk{KSPLGMRES} and \pk{KSPDGMRES}. However, neither handle variable
preconditioning, and \pk{KSPDGMRES} does not support complex arithmetic.
Moreover, extending the recycling capabilities of such methods to sequence of linear systems
with smoothly varying coefficient matrices and right-hand sides is not trivial.
On the other hand, HPDDM offers support for the
generalized conjugate residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR), which does not suffer from the aforementioned
limitations.

@@

For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}s:
\begin{itemize}
    \item \pk{MATAIJ}: standard sequential or parallel sparse matrix, based on compressed sparse row format;
    \item \pk{MATSEQBAIJ}: block sparse matrix, based on block compressed sparse row format;
    \item \pk{MATSEQSBAIJ}: symmetric block sparse matrix stored in upper triangular form;
    \item \pk{MATSHELL}: user-defined matrix;
    \item \pk{MATNEST}: block-defined matrix with nested submatrices;
    \item \pk{MATAIJCUSPARSE}: sequential or parallel sparse matrix, offloaded to a GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}
The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    %\item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: embedded Krylov method;
    \item \pk{PCMAT}: matrix multiplication;
    %\item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    %\item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}: hierarchical matrices~\cite{10.1145/3232850,ambartsumyan2020hierarchical}; 
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like
preconditioners~(\cref{it:hpddm,it:asm,it:bjacobi}) is that they most often
rely on exact or inexact factorizations~(\cref{it:lu,it:ilu}) as subdomain
solvers.  In these cases, it is possible to access the so-called factored
matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$,
%$\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to
take advantage of blocked forward eliminations and backward substitutions from
the various factorization packages interfaced with PETSc. In the case of a
sparse matrix, this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL PARDISO or
CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense matrix:
ScaLAPACK and Elemental~\cite{poulson2013elemental}.  Future work will consider
extending the multigrid framework \pk{PCMG}, as well as other preconditioning
classes to increase arithmetic intensity of the preconditioner application
phase for blocks of right-hand sides.

@@

    \item in PETSc with unpreconditioned and unrestarted GMRES($\infty$) \pk{KSPGMRES};

@@

In all the cases, convergence is declared when the initial unpreconditioned residual is reduced by 10 orders of magnitude.

@@

processes and they can be launched using the following command lines. \\[-4pt]

@@

        \caption{\# of iterations needed to converge for various

@@

        the numbers of iterations needed by \pk{KSPHPDDM} match with the respective

@@

        xlabel={\# of right-hand sides ($N$)},

@@

The mini-app is then used to benchmark intranode performance of matrix--matrix multiplications, either with multiple OpenMP threads, or by offloading the operation to a GPU and using cuSPARSE~\cite{cusparse-web-page} as interfaced in PETSc.

@@

    \subsubsection{Linear stability analysis}
In this section, we apply recycling Krylov methods in the context of linear
stability analysis and consider the solution of the following generalized eigenvalue problem:
\begin{equation}\label{eq:lsa}
    J(q_b)x = \lambda \begin{bmatrix}M & 0 \\ 0 & 0 \end{bmatrix} x,
\end{equation}
where $J$ is the Jacobian of the incompressible steady-state Navier--Stokes equation
evaluated using a given base flow $q_b = \begin{bmatrix}u_b \\ p_b\end{bmatrix}$ and
$M$ the discretization of the mass matrix on the space of velocities.
The mini-app employed for the numerical results is built on top of
FreeFEM~\cite{hecht2012new} and it is available at
\url{https://github.com/prj-/moulin2019al}.
Interested
readers are referred to~\cite{moulin2018al} for more details and for larger
runs.

@@

The Krylov--Schur algorithm converges in six outer iterations (restarts), with a total of 46 inner solves~\cref{eq:ks}.
Recycling of Krylov subspaces with \pk{KSPHPDDM} provides algorithmic speedup by lowering the number of iterations per inner solve,
as reported in~\cref{fig:comparison-slepc}, from 80 with \pk{KSPFGMRES} to 45.
A similar speedup can be observed in terms of runtime, with the time spent for the solution of the eigenvalue problem being reduced from \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1509}{\sec} to \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1052}{\sec}. These timings, and all that follow, have been obtained on {Irène}, a system composed of \pgfmathprintnumber[assume math mode=true]{1656} nodes with two 24-core Intel Xeon Platinum 8168 clocked at \SI{2.7}{\giga\hertz}.
\pgfplotstableread{
sys hpddm fgmres
1   77    77
2   54    83
3   53    83
4   52    81
5   52    81
6   51    78
7   51    79
8   51    80
9   51    80
10  51    77
11  51    79
12  51    79
13  51    79
14  53    77
15  52    79
16  49    78
17  51    77
18  51    79
19  52    79
20  52    79
21  51    77
22  51    78
23  51    79
24  53    80
25  54    79
26  52    79
27  49    75
28  51    78
29  51    78
30  51    77
31  50    78
32  55    81
33  56    79
34  54    80
35  53    80
36  53    78
37  50    78
38  52    79
39  53    77
40  52    77
41  54    80
42  53    80
43  55    80
44  53    79
45  52    77
46  53    80
}\loadedtableSLEPc
    \begin{figure}[!t]
\centering \begin{tikzpicture}
    \begin{axis}[xlabel={Krylov--Schur system index},ylabel={\# of
        iterations},xmin=1,xmax=46,ymin=0,ymajorgrids,
        legend columns=1,enlarge x limits=0.05, height=5.5cm,
        legend style={font=\small,at={(0.98,0.02)}, legend cell align=left,
        anchor=south east}]
\addplot[very thick,last]     table[x expr=\thisrowno{0},y expr=\thisrowno{2}, col sep=space] {\loadedtableSLEPc};\label{pgfplots:m0}
\addplot[very thick,myred,densely dashed]       table[x expr=\thisrowno{0},y expr=\thisrowno{1}, col sep=space] {\loadedtableSLEPc};\label{pgfplots:m1}
\legend{{\pk{KSPFGMRES}},{\pk{KSPHPDDM}}}
\end{axis}
\end{tikzpicture}
        \caption{\# of inner iterations for systems~\cref{eq:ks} with~\ref{pgfplots:m1} or without~\ref{pgfplots:m0} recycling.\label{fig:comparison-slepc}}
    \end{figure}

\subsubsection{Blocking inside SLEPc\label{sec:lobpcg}}
The implementations of the Locally Optimal Block Preconditioned Conjugate
Gradient~\cite{knyazev2001toward}~(LOBPCG) and the Contour Integral Spectrum
Slicing method~\cite{sakurai2003projection}~(CISS) in SLEPc were previously not
using blocking when applying the preconditioner or solving linear systems. In
version 3.14.0, these solvers have been adapted to employ \pk{KSPMatSolve} and
\pk{PCMatApply}.  In the case of LOBPCG, which is a purely blocked method, all
steps were already implemented as block operations, except the applications of
the preconditioner, which undermined the benefits of blocking.
%With the new
%implementation, the gains are striking, as reported in~\cref{sec:lobpcg}.
The performance of LOBPCG implemented in SLEPc as \pk{EPSLOBPCG} is studied using three different preconditioners:
\begin{itemize}
    \item \pk{PCASM}: one-level overlapping Schwarz method with one level of overlap and exact Cholesky factorizations in each subdomain;
    \item \pk{PCHPDDM}: multilevel overlapping Schwarz method, see~\cref{sec:PCHPDDM};
    \item \pk{PCGAMG}: algebraic multigrid method~\cite{adams2004ultra}.
\end{itemize}
More details on these solvers are given in the next section where focus is put
on preconditioning rather than Krylov methods.  We note here that only
\pk{PCASM} and \pk{PCHPDDM} currently handle blocking, see the
list~\cref{sec:interface-ksp}, and that \pk{PCHPDDM} and \pk{PCGAMG} have lower
contraction factors than the cheaper alternative \pk{PCASM}.

@@

The script available at \url{https://github.com/prj-/jolivet2020petsc} may be
used to solve the same eigenproblem using \pk{EPSCISS} by providing the
additional command line argument \pk{-eps\_type ciss}.
However, this eigensolver is mostly suited for finding interior eigenpairs.
For fairness, its performance is not reported, but the options from the script
can be readily used for problems where \pk{EPSCISS} is an appropriate solver.
We finally note that blocking support is also available and tested in SLEPc for \pk{EPSSUBSPACE}
and \pk{NEPCISS} since version 3.14.0.

@@

HPDDM implements the same one-level overlapping Schwarz methods as PETSc as
well as optimized Schwarz methods~\cite{gander2006optimized}, which may be used
in PETSc with \pk{PCSetModifySubMatrices}, see~\cite{gander2019new}.  It also provides support for
defining robust two-level methods using the Generalized Eigenvalue problem on
the Overlap (GenEO) framework, for finite
elements~\cite{spillane2011robust,spillane2013abstract} and boundary
elements~\cite{marchand2019two}. Results in this work present the first
high-level access to GenEO from PETSc and its composable solver
infrastructure~\cite{brown2012composable}. We note that GenEO has been
recently implemented in other libraries as well~\cite{butler2020high}.

@@

conditions. In the case of the 1D Poisson

@@

\caption{Notations and basics of overlapping Schwarz methods.\label{fig:poisson}}

@@

For performance reasons, it is better to compute the same number of
eigenvectors on each process, since in this case, the coarse operator will be
assembled using symmetric or general blocked matrix formats. With
usually large block sizes, this has consequent impacts on memory, floating-point, and message passing performances.
% When applying the preconditioner,
%\texttt{MPI\_Gather} and \texttt{MPI\_Scatter} operations are used to move
%between levels, instead of the variable-size alternatives \texttt{MPI\_Gatherv}
%and \texttt{MPI\_Scatterv}, that could exhibit worse performance.
In addition,
standard eigensolvers such as \pk{EPSKRYLOVSCHUR} from SLEPc, cannot cheaply
evaluate the exact number of eigenpairs in a given interval of $\mathbb{R}^+$.
Instead, it is often more efficient to provide an upper bound of the number of
eigenpairs.

@@

decomposition methods, \pk{PCNN} and \pk{PCBDDC} must be supplied a \pk{MatIS}

@@

it is possible to get the Dirichlet operators

@@

overlapping Schwarz methods are known for converging slowly when there is in
fact no overlap. The proposed methodology is not strictly equivalent to the

@@

	\varepsilon_{ij}(u) = \frac{1}{2} \left(\frac{\partial u_i}{\partial x_i} + \frac{\partial u_j}{\partial x_j} \right), \mu = \frac{E}{2(1+\nu)}, \text{ and }\lambda = \frac{E\nu}{1-2\nu}.

@@

    \subfloat[\parbox{5cm}{\# of inner coarse iterations for various values of $M$, the number of second-level subdomains.}\label{fig:inexact}]{

@@

configuration with 256 second-level subdomains~\ref{pgfplots:p-256}, requires 64 seconds to setup and 31 for the solution of the linear system.
